---
title: "🌿 NDVI over Oʻahu with Sentinel-2 (R Tutorial)"
subtitle: "From Copernicus Browser to NDVI map in minutes"
author: "Your Name"
format:
  html:
    theme: cosmo
    toc: true
    toc-title: "On this page"
    toc-depth: 3
execute:
  echo: true
  message: false
  warning: false
editor: visual
---

```{r}
library(here)
library(terra)
library(sf)
```

## NDVI (Normalized Difference Vegetation Index)

NDVI is a simple, per-pixel indicator of “greenness” (photosynthetic activity).

### What it measures

Plants reflect more **near-infrared (NIR)** light and absorb **red** light. NDVI combines those two bands to highlight living vegetation.

### Formula

$$
\mathrm{NDVI} \;=\; \frac{\text{NIR} - \text{Red}}{\text{NIR} + \text{Red}}
$$

```{=html}
<!-- Plain-text fallback (if math is disabled):
NDVI = (NIR - Red) / (NIR + Red)
-->
```

### Typical value ranges

-   **0.6–0.9**: dense, healthy vegetation\
-   **0.2–0.5**: sparse/brown vegetation, shrubs/grass\
-   **0–0.2**: bare soil / urban\
-   **\< 0**: water, clouds, shadows, snow

### What data do we need to calculate NDVI

-   We can use the [Copernicus Open Access Hub](https://browser.dataspace.copernicus.eu/) to grab data form the satellite Sentinel 2

-   We specifically need Band 8 data for near-infared and Band 4 data for red

```{r}
# Paths to your .jp2 files
red_path <- here("data/band_data/T04QFJ_20250906T210919_B04_10m.jp2")
nir_path <- here("data/band_data/T04QFJ_20250906T210919_B08_10m.jp2")
```

### These .jp2 files contain raster data

-   Raster data is a way to represent geography as a **grid of cells (pixels)**. Each cell has a single value (e.g., elevation, temperature, NDVI) and the whole grid is georeferenced so every cell aligns to a real-world location

-   The `terra` package helps us read raster data into R

```{r}
# Read bands directly
red <- rast(red_path)
nir <- rast(nir_path)
```

### Calculating NDVI

$$
\mathrm{NDVI} \;=\; \frac{\text{NIR} - \text{Red}}{\text{NIR} + \text{Red}}
$$

```{r}
ndvi <- (nir - red) / (nir + red)
```

### Plotting NDVI

```{r}
plot(ndvi, main = "NDVI")
```

### Trimming the plot:

```{r}
# Create extent in lat/lon
oahu_ext_latlon <- ext(-158.3, -157.6, 21.2, 21.7)

# Create a temporary raster with this extent in WGS84
temp_raster <- rast(oahu_ext_latlon, crs = "EPSG:4326")

# Project to match your NDVI's CRS
temp_raster_utm <- project(temp_raster, crs(ndvi))

# Get the extent in UTM coordinates
oahu_bbox <- ext(temp_raster_utm)

# Crop your NDVI
ndvi_oahu <- crop(ndvi, oahu_bbox)
plot(ndvi_oahu, main = "NDVI - Oahu Area")
```

```{r}
library(ggplot2)
library(tidyterra)

ggplot() +
  geom_spatraster(data = ndvi_oahu) +
  scale_fill_gradient2(low = "#8B4513", mid = "#FFD700", high = "#006400",
                       midpoint = 0.4,
                       name = "NDVI",
                       na.value = "transparent") +
  labs(title = "Normalized Difference Vegetation Index",
       subtitle = "Oʻahu, Hawaiʻi",
       caption = "Data: Sentinel-2, September 2025") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 12),
        legend.position = "bottom")
```

```{r}
library(mapview)

# Reduce resolution significantly (factor of 8-10 should work)
ndvi_small <- aggregate(ndvi_oahu, fact = 8, fun = mean)

# Check the new size
ncell(ndvi_small)  # Should be much smaller now

# Now map it
mapview(ndvi_small, 
        col.regions = c("#8B4513", "#FFD700", "#9ACD32", "#228B22", "#006400"),
        layer.name = "NDVI",
        alpha = 0.7)
```

```{r}
library(mapview)

# Reduce resolution
ndvi_small <- aggregate(ndvi_oahu, fact = 8, fun = mean)

# Mask out negative values (ocean/water)
ndvi_veg_only <- ndvi_small
ndvi_veg_only[ndvi_veg_only < 0] <- NA

# Create a focused color palette for 0-0.6 range
mapview(ndvi_veg_only,
        at = seq(0, 0.6, 0.1),  # Specific breaks from 0 to 0.6
        col.regions = c("#D2691E", "#DEB887", "#F0E68C", "#ADFF2F", "#32CD32", "#228B22", "#006400"),
        layer.name = "NDVI",
        alpha = 0.8,
        na.color = "#E6F3FF")  # Light blue for ocean/NA values
```

```{r}
# Create custom breaks and colors
breaks <- c(-1, 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 1)
colors <- c("#4A90E2",      # Ocean (blue)
           "#CD853F",       # Bare soil/urban (tan)
           "#DEB887",       # Very sparse (light tan)
           "#F0E68C",       # Sparse (light yellow)
           "#ADFF2F",       # Moderate (yellow-green)
           "#32CD32",       # Good vegetation (lime green)
           "#228B22",       # Dense vegetation (forest green)
           "#006400")       # Very dense (dark green)

mapview(ndvi_small,
        at = breaks,
        col.regions = colors,
        layer.name = "NDVI",
        alpha = 0.8)
```

```{r}
library(viridis)

# Custom function to handle ocean vs land
create_ndvi_palette <- function(n = 100) {
  ocean_color <- "#2E86AB"  # Nice blue for ocean
  land_colors <- viridis(n, option = "D")  # Green-focused palette for land
  c(ocean_color, land_colors)
}

mapview(ndvi_small,
        at = c(-1, 0, seq(0.05, 0.6, length.out = 12)),
        col.regions = create_ndvi_palette(12),
        layer.name = "NDVI",
        alpha = 0.8)
```

```{r}
library(mapview)

# Reduce resolution
ndvi_small <- aggregate(ndvi_oahu, fact = 8, fun = mean)

# Define breaks based on your categories
breaks <- c(-1, 0, 0.2, 0.5, 0.6, 1)

# Colors matching the categories
colors <- c("#4A90E2",      # < 0: Water (blue)
           "#8B4513",       # 0-0.2: Bare soil/urban (brown)
           "#CD853F",       # 0.2-0.5: Sparse/brown vegetation (tan/light brown)
           "#228B22",       # 0.5-0.6: Transition to dense (forest green)
           "#006400")       # 0.6+: Dense, healthy vegetation (dark green)

mapview(ndvi_small,
        at = breaks,
        col.regions = colors,
        layer.name = "NDVI",
        alpha = 0.8)
```

```{r}
# Use the higher resolution version for more accurate statistics
ndvi_values <- values(ndvi_oahu, na.rm = TRUE)

# Define the same categories
categories <- cut(ndvi_values, 
                  breaks = c(-Inf, 0, 0.2, 0.5, 0.6, Inf),
                  labels = c("Water/Clouds", "Bare/Urban", "Sparse Vegetation", 
                            "Moderate Vegetation", "Dense Vegetation"),
                  include.lowest = TRUE)

# Calculate counts and proportions
category_counts <- table(categories)
category_props <- prop.table(category_counts) * 100

# Create a nice summary
results <- data.frame(
  Category = names(category_counts),
  Pixel_Count = as.numeric(category_counts),
  Percentage = round(as.numeric(category_props), 2)
)

print(results)
```

```{r}
# For land-only analysis (excluding water/negative values)
land_values <- ndvi_values[ndvi_values >= 0]

land_categories <- cut(land_values,
                      breaks = c(0, 0.2, 0.5, 0.6, 1),
                      labels = c("Bare/Urban", "Sparse Vegetation", 
                                "Moderate Vegetation", "Dense Vegetation"),
                      include.lowest = TRUE)

land_counts <- table(land_categories)
land_props <- prop.table(land_counts) * 100

land_results <- data.frame(
  Category = names(land_counts),
  Pixel_Count = as.numeric(land_counts),
  Percentage_of_Land = round(as.numeric(land_props), 2)
)

print("Land cover analysis (excluding water):")
print(land_results)
```

```{r}
library(ggplot2)

# Create a bar chart
ggplot(results, aes(x = Category, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Water/Clouds" = "#4A90E2",
                              "Bare/Urban" = "#8B4513",
                              "Sparse Vegetation" = "#CD853F",
                              "Moderate Vegetation" = "#BDB76B",
                              "Dense Vegetation" = "#006400")) +
  labs(title = "Land Cover Distribution - Oʻahu",
       subtitle = "Based on NDVI Categories",
       x = "Category", y = "Percentage of Total Area") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  geom_text(aes(label = paste0(Percentage, "%")), 
            vjust = -0.3, size = 3)
```

```{r}
# Quick summary stats
cat("NDVI Summary for Oʻahu:\n")
cat("Total pixels analyzed:", length(ndvi_values), "\n")
cat("Land pixels (NDVI ≥ 0):", length(land_values), "\n")
cat("Water/ocean pixels:", sum(ndvi_values < 0), "\n\n")

cat("Mean NDVI (land only):", round(mean(land_values), 3), "\n")
cat("Median NDVI (land only):", round(median(land_values), 3), "\n")
cat("Max vegetation index:", round(max(ndvi_values), 3), "\n")
```
